\documentclass[12pt]{article}

\title{Tic-Tac-Toe with SAT}
\author{William John Holden}
\date{\today}

\usepackage{amsmath,amsthm}
\usepackage{mathtools}
%\usepackage{parskip}
\usepackage{amssymb}
\usepackage[margin=1.0in]{geometry}
\usepackage[hyphens]{url}
\usepackage[hidelinks]{hyperref}
\usepackage{listings}
\lstset{basicstyle=\ttfamily,columns=flexible,frame=single,breaklines=true,linewidth=1\textwidth}

\newcommand{\tictactoe}[9]{\begin{center}\begin{tabular}{ c | c | c }{$#1$} & {$#2$} & {$#3$} \\ \hline {$#4$} & {$#5$} & {$#6$} \\ \hline {$#7$} & {$#8$} & {$#9$} \end{tabular}\end{center}}

\begin{document}

\maketitle

\section{Introduction}

In this exercise we will construct a series of boolean satisfiability clauses to find a winning move for the game Tic-Tac-Toe.
Throughout this exercise, assume that we are playing as $x$ and that it is our turn to play for any game shown.
We treat the SAT solver as a ``black box'' and will not worry about how it works.
A SAT solver accepts a \textit{formula} of \textit{clauses} containing boolean \textit{variables} in \textit{conjunctive normal form} (CNF).
An example of a formula in CNF is

\begin{equation}(A \vee \neg B)(C)(C \vee D \vee E)(B \vee \neg C).\end{equation}

This formula is \textit{satisfiable} with the literals $A=T$, $B=T$, $C=T$, $D=F$, and $E=F$.
This is not the only satisfying assignment.
$D$ and $F$ may each take the values $T$ or $F$, but we are constrained in $C=T$.
If $C=F$ then the clause $(C)$ cannot be satisfied, and any unsatisfied clause falsifies the entire formula.

Some formulas cannot be satisfied under any assignment of literals to variables. An obvious example is $(A)(\neg A)$.
A less obvious example is

\begin{equation}(A \vee \neg B)(B \vee \neg C)(C)(\neg A \vee \neg B \vee \neg C).\end{equation}

In this case, the $(C)$ clause implies $C=T$, which means $B=T$, which then implies $A=T$, but now the final clause is $(F \vee F \vee F)$ which falsifies the formula.

A \textit{SAT solver} is a computer program for finding a set of satisfying literals or proving if no such satisfying assignment exists.
Though the boolean satisfiability problem is NP-complete, state of the art SAT solvers are very efficient in practice.
Martin Horenovsky has a very nice article at \url{https://codingnest.com/modern-sat-solvers-fast-neat-underused-part-1-of-n/} on this subject.
I originally learned of the satisfiability problem and SAT solvers attending a course on NP-Complete Problems from Alexander Kulikov (\url{https://www.edx.org/course/np-complete-problems-uc-san-diegox-algs203x}).

Others have successfully solved Sudoku puzzles using SAT solvers.
In this exercise, we will do the same for tic-tac-toe.
Of course, tic-tac-toe is much less computationally demanding than Sudoku.
Tic-tac-toe has a smaller board ($3 \times 3$ instead of $9 \times 9$) and it has fewer constraints.
If we consider the size of the tic-tac-toe puzzle to be $n$ for dimensions $\sqrt{n} \textrm{ rows} \times \sqrt{n} \textrm{ columns} = x \times y$ then we can iterate over every row and column to find a winning move in

\begin{align}O(x \times y) + O(y \times x ) + O(2 \textrm{ diagonals} \times \sqrt{n}) &= \\
O(xy)+O(xy)+O(2\sqrt{n})
&=O(n).
\end{align}

Using an exponential time tool for a linear time problem may seem crazy, but the goal here is to learn to \textit{reduce} problems to SAT.

\section{Reduction to SAT}

\subsection{Variables}

We need to express the positions of our tic-tac-toe board as boolean variables. This is pretty easy. We assign a variable letter to each position in reading order:

\tictactoe{a}{b}{c}{d}{e}{f}{g}{h}{i}

There are actually \textit{three} possible states for each position: X, O, and unassigned. However, in this problem we do not need to concern ourselves with the third state.
We will consider a variable to be ``true'' if it is marked wtih X and false otherwise.

\subsection{Contraint 1: Winning Combinations}

There are eight total winning combinations in tic-tac-toe: satisfy any of the three rows, any of the columns, or either of the diagonals.
I should note that the word \textit{or} in the previous sentence is an inclusive or; winning combinations allow for both a row and a column, both a row and a diagonal, and both a column and a diagonal. For example, the assignment

\tictactoe{o}{o}{x}{o}{o}{x}{x}{x}{}

can be won by setting $i=x$.  We can express the eight winning combinations as

\begin{equation}(a \wedge b \wedge c) \vee (d \wedge e \wedge f) \vee (g \wedge h \wedge i)
\vee
(a \wedge d \wedge g) \vee (b \wedge e \wedge h) \vee (c \wedge f \wedge i)
\vee
(a \wedge e \wedge i) \vee (c \wedge e \wedge g).\end{equation}

This long formula is in \textit{disjunctive normal form} (DNF) and not the CNF we want.
We could use rules of discrete mathematics, such as DeMorgan's Law and Distributivity, but in this paper on computational tools let us instead use \textit{Mathematica's} \texttt{BooleanConvert} function (\url{https://reference.wolfram.com/language/ref/BooleanConvert.html}).
\texttt{BooleanConvert} accepts an expression and converts it to the desired form.
(\textit{Mathematica} functionality is available at \url{https://www.wolframalpha.com}).

\begin{lstlisting}
In[1] := BooleanConvert[((a && b && c) || (d && e && f) || (g && h && i) || (a && d && g) || (b && e && h) || (c && f && i) || (a && e && i) || (c && e && f)),"CNF"]
Out[1] := (a || b || c || d || g) && (a || b || f || g) && (a || b || f ||  i) && (a || c || d || h) && (a || c || e || h) && (a || e || f || g) && (a || e || i) && (a || f || h) && (b || c || d || i) && (b || d || e || f || h) && (b || d || e || i) && (b || d || f ||  i) && (b || e || f || g) && (b || e || g || i) && (b || f || g || i) && (c || d || e || h) && (c || d || e || i) && (c || d || h || i) && (c || e || g) && (c || f || g || h || i)
\end{lstlisting}

Now we have our first set of constraints that we can provide to a SAT solver.
An assignment of literals to the variables $a$ through $i$ is a winning move if all three variables in any row, column, or diagonal are $x$.

\subsection{Contraint 2: Board Configuration}

The game will be shown with some configuration of $x$ and $o$ settings on the board.
For each position with an $x$ we will create a clause of unit length where that variable must be true.
All positions marked $o$ induce a clause where that variable must be false.
For example, given the game

\tictactoe{x}{o}{x}{o}{x}{o}{o}{}{}

we will set $a=T$, $b=F$, $c=T$, $d=F$, $e=T$, $f=F$, $g=F$ with the clauses

\begin{equation}(a)(\neg b)(c)(\neg d)(e)(\neg f)(\neg g).\end{equation}

These clauses must be generated based on game state.

\subsection{Constraint 3: One Move}

We assume that it is our turn to play, and we only get to make one move.
For this, we use the \textit{exclusive or} operation against all of the unset variables.
In the game

\tictactoe{x}{}{x}{}{o}{}{o}{}{o}

we create constraint clauses to show $b \oplus d \oplus f \oplus h$.
We can punch this into \textit{Mathematica} if we want to, but we will need to generalize nested $\oplus$ operations.

Exclusive or is defined as either but not both of two propositions. Using DeMorgan's Law, $\neg (x \wedge y) = \neg x \vee \neg y$, we can easily convert this statement into CNF.

\begin{equation}A \oplus B = (A \vee B) \wedge \neg (A \wedge B) = (A \vee B) \wedge (\neg A \vee \neg B).\end{equation}

If $B = C \oplus D$ then $A \oplus B = A \oplus (C \oplus D) = A \oplus ((C \vee D) \wedge (\neg C \wedge \neg D))$. 

\end{document}
